The state learner \\ {-- a super learner for right-censored data}

### 2024-05-08 tex/jrss-version/state-learner-jrss-style.tex
progress

% ANDERS / THOMAS
We here consider the situation where $\Lambda_1$, $\Lambda_2$, and $\Gamma$ are
nuisance parameters and an the estimator of interest is obtained from estimators
of these nuisance parameters. An example is an estimator of the cumulative
incidence curve, which can be obtained from estimators of $\Lambda_1$ and
$\Lambda_2$.
% THOMAS: Jeg ville så også tilføje: We give another example in Section 7.

@@

In this section we report results from a simulation study without competing
risks. In the next section, we consider a setting with competing risks. Here we
first compare the state learner to an IPCW based discrete super learner that
uses either the Kaplan-Meier estimator or a Cox model to estimate the censoring
probability. We next compare the state learner to the super learner proposed by
\cite{westling2021inference}.
% ANDERS: kan du skrive her hvad forskellen er mellem og formaalet er
% med de to simulationsstudierscenarier?
% i den ene sammenligner vi med IPCW i den anden med westling, ik?
% Maaske en figur (flowchart) eller tabel kunne hjaelpe laeseren at faa overblik.
% Maaske separation i to subsections kunne hjaelpe yderliger
%
% THOMAS: Jeg ved ikke, om jeg ville sige at der er to simulationstudier, der er
% bare to ting vi undersøge med det samme studie. Data er genereret på samme
% måde til at undersøge de to spørgsmål. Har opdateret introen til afsnittet, er
% det bedre?

We generate data according to a distribution motivated from a real
dataset in which censoring depends on the baseline covariates. We
simulate data based on a prostate cancer study described in
\citep{kattan2000pretreatment}. The outcome of interest was the time
to tumor recurrence, and five baseline covariates were used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
(HT, yes/no) and clinical stage (CS, six values). The study was
designed such that a patient's radiation dose depended on when the
patient entered the study \citep{gerds2013estimating}. This in turn
implied that the time of censoring depended on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a
sensitivity analysis was conducted based on simulated data. We use the
same simulation setup, where event and censoring times are generated
according to parametric Cox-Weibull models estimated from the original
data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original
data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
simulation setting as ``dependent censoring''. We also considered a
simulation setting where data were generated in the same way, except
that censoring was generated completely independently. We refer to
this simulation setting as ``independent censoring''.

@@

We compare the state learner to two IPCW based super learners: The
first superlearner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
% ANDERS: det er uklar for mig hvordan
% de super learners valger/kombinerer de learners der ingaar i library: er det
% Brier score? over time?
% 
% THOMAS: De er diskrete super learner baseret på Score function med summary
% measure 'ibs', så det er integrated brier score.
Each super learner provides a learner for the cumulative hazard
function for the outcome of interest. From the cumulative hazard
function a risk prediction model can be obtained (c.f.,
equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
performance of each super learner by calculating the index of
prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
horizon (36 months) for the risk prediction model provided by the
super learner. The IPA is 1 minus the ratio between the model's Brier
score and the null model's Brier score, where the null model is the
model that does not use any covariate information. The IPA is
approximated using a large (\( n = 20,000 \)) independent data set of
uncensored data. As a benchmark we calculate the performance of the
risk prediction model chosen by the oracle selector, which uses the
large data set of uncensored event times to select the learner with
the highest IPA.

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
    % ANDERS: er det korrekt at det er averages across 1000 data sets?
    %
    % THOMAS: Ja. ...?
  }
  \label{fig:ipcw-fail}
\end{figure}

>>>>

In this section we demonstrate a theoretical property of the state
learner which is useful for targeted learning (c.f., Section
\ref{sec:targeted-learning}). Specifically we consider an estimator of
a target parameter which is obtained by substituting the state learner
estimates of the nuisance parameters $\Lambda_1$, $\Lambda_2$, and
$\Gamma$. An example is an estimator of the cumulative incidence
curve, which can be obtained from estimators of $\Lambda_1$ and
$\Lambda_2$. Another example is provided in Section \ref{sec:targeted-learning}.

@@

In this section we report results from a simulation study. In the
first part we consider estimation of the conditional survival
function. Here we compare the state learner to two IPCW based discrete
super learners that use either the Kaplan-Meier estimator or a Cox
model to estimate the censoring probability
\cite{gonzalez2021stacked}.  In the second part we compare the state
learner to the super learner proposed by \cite{westling2021inference}.

In both parts we use the same data generating mechanism. We generate
data according to a distribution motivated from a real dataset in
which censoring depends on the baseline covariates. We simulate data
based on a prostate cancer study described in
\citep{kattan2000pretreatment}. The outcome of interest was the time
to tumor recurrence, and five baseline covariates were used to predict
outcome: prostate-specific antigen (PSA, ng/mL), Gleason score sum
(GSS, values between 6 and 10), radiation dose (RD), hormone therapy
(HT, yes/no) and clinical stage (CS, six values). The study was
designed such that a patient's radiation dose depended on when the
patient entered the study \citep{gerds2013estimating}. This in turn
implied that the time of censoring depended on the radiation dose. The
data were re-analyzed in \citep{gerds2013estimating} where a
sensitivity analysis was conducted based on simulated data. We use the
same simulation setup, where event and censoring times are generated
according to parametric Cox-Weibull models estimated from the original
data, and the covariates are generated according to either marginal
Gaussian normal or binomial distributions estimated from the original
data \citep[c.f.,][Section~4.6]{gerds2013estimating}. We refer to this
simulation setting as ``dependent censoring''. We also considered a
simulation setting where data were generated in the same way, except
that censoring was generated completely independently. We refer to
this simulation setting as ``independent censoring''.

@@

We compare the state learner to two IPCW based super learners: The
first superlearner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
% ANDERS: describe that you use a single evaluation time point and say how many censored, event happen before it.
Each super learner provides a learner for the cumulative hazard
function for the outcome of interest. From the cumulative hazard
function a risk prediction model can be obtained (c.f.,
equation~(\ref{eq:surv-def}) with $\Lambda_2 = 0$). We measure the
performance of each super learner by calculating the index of
prediction accuracy (IPA) \citep{kattan2018index} at a fixed time
horizon (36 months) for the risk prediction model provided by the
super learner. The IPA is 1 minus the ratio between the model's Brier
score and the null model's Brier score, where the null model is the
model that does not use any covariate information. The IPA is
approximated using a large (\( n = 20,000 \)) independent data set of
uncensored data. As a benchmark we calculate the performance of the
risk prediction model chosen by the oracle selector, which uses the
large data set of uncensored event times to select the learner with
the highest IPA.

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}
